---
title: "PA1_template"
author: "Marta Fisher"
date: "March 25, 2016"
output: html_document
---
##Loading and preprocessing the data
1.Read the data and show the first few rows of data
```{r}
StepsData <- read.csv("C:/Users/fisherm/Documents/Coursera/reproducible_research/activity.csv")
head(StepsData)
```

2.Process or transform the data as needed.
We do need to load 2 packages
and change the date column to date format
```{r, results='hide'}
library(dplyr)
library(ggplot2)
StepsData <- mutate(StepsData, date = as.Date(date))
```

##Mean total number of steps per day? 
(ignore missing values)
1. Calculate the total number of steps taken per day
```{r}
sum(StepsData[, 1], na.rm = TRUE)
```
2. Make a histogram of the total number of steps taken each day
```{r}
baseplot <- ggplot(StepsData, aes(date))
baseplot + geom_bar(aes(weight = steps)) + labs(y = "total number of steps", x = "Day in 2012", title = "Number of steps by day") + scale_x_date(date_breaks = "1 week", date_labels = "%m/%d")
```
3. Calculate and report the mean and median of the total number of steps/day  

Mean steps/day
```{r}
summarize(group_by(StepsData, date), mean(steps))
```
Median steps/day
```{r}
summarize(group_by(StepsData, date), median(steps))
```

##Average Daily Activity Pattern?
1. Make a time series plot of the 5 min interval and the avg number of steps taken, averaged across all days.
2. Which 5 min interval, on average across all days, contains the maximum number of steps?

##Imputing missing values
1. Calculate and report the total # of missing values.
2. Devise a strategy for filling in all the missing values in the dataset.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps/day. Do the values differ from the 1st part of the assigment? What is the impact of imputing missing data on teh estimates of total daily steps?

##Are there differences in weekeday and weekend activity patterns?
1. Create a new factor variable in the dataset with 2 levels, Weekday and Weekend.
2. Make a panel plot containing a times series plot of the 5 min interval and the avg # of steps taken, averaged across weekday vs weekend.

